---
title: "Reproducable Research Assignment  (Week 2)"
output: html_document
---

This is a project on a dataset of `moement sensors in wearable devices'. The dataset loaded contained the number of steps in 5-min intervals for a couple of days. The variablesa re the number of steps, date, and intervals. 

In this project we compare the mean of the number of steps throughout a typical day. Also we try to impute data by replacing *NA* values with some reasonable values. We compare results for different days of weeks. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

## Read files

In this part, we read the source file from web. Next we unzip the zip file, and read the csv file into *df* var. 
 

```{r readfile}
link1 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip" 
download.file( url = link1,destfile = "activity.zip") 
df_original <- read.csv(unzip("activity.zip"))
```

## What is mean total number of steps taken per day?
 
In this part, we remove **NA** values here.  Then we obtain the total number of steps for each day. The mean and median number of steps are also computed. 

```{r total}
df <- df_original[!is.na(df_original[,1]),]
StepDaily <-  tapply(df$steps,df$date,sum)
print("Number of steps per day:")
StepDaily
```
### mean of steps: 
```{r mean}
mean(StepDaily,na.rm = T)
```
median of steps: 
```{r med}
median(StepDaily,na.rm = T)
```
```{r plotHist}
hist(StepDaily, xlab = "number of steps per day", main = "Histogram of steps per day")
```


## What is the average daily activity pattern?

In this part, we compare the activity pattern thourghout a typical day by average of the number of steps durign 5-min intervals.

```{r time series}
library(dplyr)
TS <- df %>% group_by(interval) %>% summarise(meanSteps = mean(steps))
plot(TS$interval,TS$meanSteps,type="l",xlab = "5-min intervals during a typical day", ylab = "mean of number of steps", main = "time series plot for average steps in a day")
```

The 5-min Interval with maximum Mean Steps (mean over all days) is
```{r maxStps}  
as.numeric(TS[ TS$meanSteps == max(TS$meanSteps), 1])
```
 
## Imputing missing values
In this part, we impute data to replace *NA* values with a reasonable. We replace missing values with the average of steps for similar interval over all days. 

The total number of rows with *NA* steps is
```{r missing}
sum(is.na(df_original$steps))
```
 

Here we create a new dataset from the original dataset by replacing *NA* values of **Steps** with the average steps for the same interval over all days
```{r missingFilling}
imputDF <- left_join(df_original,TS, by = "interval")
imputDF[is.na(imputDF[,1]),1] <- imputDF[is.na(imputDF[,1]),4]
imputDF <- imputDF[,-4]
```

### A histogram of  *Steps*,  the mean and median of steps

mean of steps:
```{r mean2}
impSumm <- imputDF %>% group_by(date) %>% summarise(SumStp = sum(steps))
mean(impSumm$SumStp)
```
median of steps: 
```{r med2}
median(impSumm$SumStp )
```
```{r plotHist2}
hist(impSumm$SumStp, xlab = "number of steps per day", main = "Histogram of steps per day (imputed data)")
```


## Activity patterns between weekdays and weekends

An interesting question is how the activity pattern changes according to weekend or weekday. Here are answers:

```{r week}
imputDF$weekNm <- weekdays(as.Date(imputDF$date))
weekD <- c("Tuesday","Monday","Friday","Wednesday","Thursday")
weekend <- c("Saturday","Sunday")
imputDF[imputDF$weekNm %in% weekD,5] <- "weekday"
imputDF[imputDF$weekNm %in% weekend,5] <- "weekend"
imputDF <- imputDF[,-4]
names(imputDF) <- c("steps","date","interval","day_week")
```

## Time series plot of the average number of steps across all weekdays or weekends 

```{r plotWeek}
dfAve <- imputDF %>% group_by(interval,day_week) %>% summarise(meanSteps = mean(steps))
par(mfrow = c(1, 2))
library(ggplot2) 
qplot( interval, meanSteps, facets = .~day_week,data=dfAve,geom = "line"  )
```
